---
title: "10 Tibbles Exercise Solutions"
author: "Sean Angiolillo"
date: "17 March 2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(tidyverse)
```

## 10.5 Exercises

1. How can you tell if an object is a tibble? (Hint: try printing `mtcars`, which is a regular data frame).

Printing a dataframe like `mtcars` to the console prints the entire dataframe, whereas printing a tibble gives only the first 10 rows with the classes of each column on top.

```{r}
class(mtcars)
class(as_tibble(mtcars))
```

2. Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?

We can test for differences by turning `df` into a tibble and running the same operations.

```{r eval=FALSE}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]

tbl <- as_tibble(df)
tbl$x
tbl[, "xyz"]
tbl[, c("abc", "xyz")]
```

Dataframes partially match column names so `df$x` returns the `xyz` column as a vector, whereas `tbl$x` will throw an error for an unknown column. Because it only selects one column, `df[, "xyz"]` returns a vector whereas `tbl[, "xyz"]` returns a tibble. I believe `df[, c("abc", "xyz")]` and `tbl[, c("abc", "xyz")]` are equivalent with the exception of course that the first remains a dataframe while the latter is a tibble.

3. If you have the name of a variable stored in an object, e.g. `var <- "mpg"`, how can you extract the reference variable from a tibble?

You can use double bracket `[[` but not `$`. This same behavior seems to apply for both dataframes and tibbles. Note that `tbl[[var]]` returns a vector and not a tibble.

```{r}
var <- "mpg"

mtcars[[var]] # returns a vector
mtcars$var # NULL

tbl <- as_tibble(mtcars)
tbl[[var]] # returns a vector
tbl$var # unknown column
```


4. Practice referring to non-syntactic names in the following data frame by:

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
```

* Extracting the variable called 1.

```{r}
annoying$`1` # returns a vector
```

* Plotting a scatterplot of 1 vs 2.

```{r}
ggplot(annoying, aes(x = `1`, y = `2`)) + geom_point()
```


* Creating a new column called 3 which is 2 divided by 1.

```{r}
annoying %>% mutate(`3` = `2` / `1`)
```

* Renaming the columns to one, two and three.

```{r}
annoying %>% 
    mutate(`3` = `2` / `1`) %>%
    rename(one = `1`, two = `2`, three = `3`)
```

5. What does `tibble::enframe()` do? When might you use it?

`enframe()` converts named atomic vectors or lists to two-column data frames. For unnamed vectors, the natural sequence is used as name column.

```{r}
enframe(c(a = 5, b = 7))
```

6. What option controls how many additional column names are printed at the footer of a tibble?

`n_extra` controls the number of additional column names printed at the footer of a tibble.

```{r eval=FALSE}
print(nycflights13::flights) # by default gives all extra column names in footer
print(nycflights13::flights, n_extra = 2) # will give 2 additional column names in footer
```

