---
title: "12 Tidy Data Exercise Solutions"
author: "Sean Angiolillo"
date: "18 March 2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(tidyverse)
library(gridExtra)
```

## 12.2.1 Exercises

1. Using prose, describe how the variables and observations are organised in each of the sample tables.

`table1` is tidy because each variable a column, each observation a row, each value its own cell.
```{r}
table1
```

`table2` stores variables `cases` and `population` as values of a `type` variable and creates a new variable `count` for their respective values.

```{r}
table2
```

`table3` stores the variables `cases` and `population` as one column separated with a "/" called `rate`.

```{r}
table3
```

`table4` stores the values of the `year` variable as columns filled with `cases` values in `table 4a` and `population` values in `table4b`.

```{r}
table4a
table4b
```


2. Compute the `rate` for `table2`, and `table4a + table4b`. You will need to perform four operations:

* Extract the number of TB cases per country per year.

* Extract the matching population per country per year.

* Divide cases by population, and multiply by 10000.

* Store back in the appropriate place.

```{r}
cases <- table2 %>% filter(type == "cases") %>% .$count
population <- table2 %>% filter(type == "population") %>% .$count
year <- table2 %>% filter(type == "cases") %>% .$year
country <- table2 %>% filter(type == "cases") %>% .$country

tab2_tidy <- tibble(country, year, rate = cases / population * 10000)
tab2_tidy
```

```{r}
country_b <- rep(table4a %>% .$country, each = 2)
year_b <- rep(names(table4b)[-1], 3)
cases_b <- c(rbind(table4a %>% .$`1999`, table4a %>% .$`2000`))
pop_b <- c(rbind(table4b %>% .$`1999`, table4b %>% .$`2000`))

tab4_tidy <- tibble(country_b, year_b, rate = cases_b / pop_b * 10000)
tab4_tidy
```

Which representation is easiest to work with? Which is hardest? Why?

Both can be done but perhaps 4a and 4b are more difficult to work with because values for a single variable are split into different columns. There are better ways to do this however as this chapter will show.

3. Recreate the plot showing change in cases over time using `table2` instead of `table1`. What do you need to do first?

The first step is to filter for `cases`. Then the `count` variable contains the `case` values we want to plot for each country and year.

```{r}
grid.arrange(
    ggplot(table1, aes(year, cases)) + 
        geom_line(aes(group = country), colour = "grey50") + 
        geom_point(aes(colour = country)),
    table2 %>% 
        filter(type == "cases") %>%
        ggplot(aes(year, count)) + 
            geom_line(aes(group = country), colour = "grey50") + 
            geom_point(aes(colour = country)),
    ncol = 2
)
```

## 12.3.3 Exercises

1. Why are `gather()` and `spread()` not perfectly symmetrical?
Carefully consider the following example:

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`)
```

(Hint: look at the variable types and think about column names.)

The `year` variable is intiially numeric. When the values of `year` are spread into variable names, the numeric years have been coerced to character strings. When those character years are then gathered into values of the variable `year` once more, they retain their character class. Tbe class attribute was lost.

Both `spread()` and `gather()` have a convert argument. What does it do?

The `convert` argument, if set to `TRUE`, can address the type coercion problem in the above example. Note how the year variable retains its numeric class with `convert` set to `TRUE`.

```{r}
stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`, convert = TRUE)
```

2. Why does this code fail?

```{r eval=FALSE}
table4a %>% 
  gather(1999, 2000, key = "year", value = "cases")
```

It fails because `1999` and `2000` are non-syntactic. Putting them in backticks makes the code run as intended.

```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```


3. Why does spreading this tibble fail? How could you add a new column to fix the problem?

```{r}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
people
```

Spreading fails because there are two values for a single variable (`age`) of one observation (Phillip Woods). If we added a variable representing the appearance count of an age value, then we could spread it.

```{r}
people %>%
    mutate(obs = c(1,1,2,1,1)) %>%
    spread(key, value)
```

4. Tidy the simple tibble below. Do you need to spread or gather it? What are the variables?

You need to gather it because `male` and `female` are values of the key `gender`. The variables are the logical `pregnant` the categorical `sex` and the numeric `obs`.
```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
preg %>%
    gather(key = sex, value = obs, male, female)
```

## 12.4.3 Exercises

1. What do the `extra` and `fill` arguments do in `separate()`? Experiment with the various options for the following two toy datasets.

```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")
```

In the first example, we get a warning that a piece has been discarded. The second row had four observations despite there being only three columns. We get the warning because the argument `extra` is set to "warn" by default. Changing it to "drop" would remove the warning and produce the same result. Setting it to "merge" keep the extra value in the same cell as last column in that row.

```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))
```

In the second example we get a warning that a missing piece has been filled with `NA`. This is because the `fill` argument is set to "warn" by default. Changing it to "left" or "right" removes the warning and specifies where to place the missing value.

2. Both `unite()` and `separate()` have a `remove` argument. What does it do? Why would you set it to FALSE?

The `remove` argument, if set to `FALSE`, will retain the input column in the output dataframe. It's useful if you wanted to compare before and after the union or separation. For example:

```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "right", remove = FALSE)
```

3. Compare and contrast `separate()` and `extract()`. Why are there three variations of separation (by position, by separator, and with groups), but only one unite?

Both `separate()` and `extract()` are meant to split one column into multiple columns. `separate()` does this by a splitting pattern or position. `extract()` needs regular expression groups. Variations of separation are needed because there are many ways to split one entity into many (position, separator, groups), but there's only one way to combine successfully.

## 12.5.1 Exercises

1. Compare and contrast the `fill` arguments to `spread()` and `complete()`.

Both `fill` arguments intend to replace missing values. `spread()`'s primary function is for spreading key-value pairs across multiple columns, but its `fill` argument can replace missing values with a specified value. For example,

```{r}
table2[[1,4]] <- NA
table2 %>%
    spread(key = type, value = count, fill = "missing")
```

`complete()` is intended to turn implicit missing values into explicit missing values. Its `fill` argument takes a named list for each variable letting you specify values instead of NA. So in the example below, it specifies 0 instead of `NA`.

```{r}
df <- tibble(
  group = c(1:2, 1),
  item_id = c(1:2, 2),
  item_name = c("a", "b", "b"),
  value1 = 1:3,
  value2 = 4:6
)
df %>% complete(group, nesting(item_id, item_name), fill = list(value1 = 0))
```

2. What does the direction argument to `fill()` do?

The direction argument of `fill()` determines the direction in which to fill missing values (the default "down" or "up"). For example:

```{r}
df <- data.frame(Month = 1:6, Year = c(2000, rep(NA, 5)))
df %>% fill(Year)

df <- data.frame(Month = 1:6, Year = c(rep(NA, 5), 2000))
df %>% fill(Year, .direction = "up")
```

## 12.6.1 Exercises

1. In this case study I set `na.rm = TRUE` just to make it easier to check that we had the correct values. Is this reasonable? Think about how missing values are represented in this dataset. Are there implicit missing values? What’s the difference between an `NA` and zero?

The presence of zeros in the case variable suggests that 0 indicates the true number of cases and `NA` means no data was collected for that observation (or not present). In this case, removing `NA` values would not remove any information-- with the exception of knowing which countries and years the WHO did not collect data (which could be interesting in its own right).

2. What happens if you neglect the `mutate()` step? (`mutate(key = stringr::str_replace(key, "newrel", "new_rel"))`)

Skipping the `mutate()` step would throw off the pattern we need in the next step when we `separate()` on "_".

3. I claimed that `iso2` and `iso3` were redundant with `country`. Confirm this claim.

No rows turn up in the manipulation below.

```{r}
who %>%
    select(country, iso2, iso3) %>%
    distinct() %>%
    group_by(country) %>%
    filter(n() > 1)
```

4. For each country, year, and sex compute the total number of cases of TB. Make an informative visualisation of the data.

```{r}
# tidied dataset
who_final <- who %>%
  gather(code, value, new_sp_m014:newrel_f65, na.rm = TRUE) %>% 
  mutate(code = stringr::str_replace(code, "newrel", "new_rel")) %>%
  separate(code, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)

who_final %>%
  group_by(country, year, sex) %>%
  filter(year >= 1995) %>% # little data before 1995
  summarise(cases = sum(value)) %>%
  unite(country_sex, country, sex, remove = FALSE) %>%
  ggplot(aes(x = year, y = cases, group = country_sex, colour = sex)) +
    geom_line()
```
